On the Gravitational stability of gravito-turbulent accretion disks

### 2016-03-24 paper/application.tex
rough revision done

PPDs become irradiation-dominated at large distances from the star,
where $T\to\tirr$ and $\theta\to 1$ \citep{dalessio97,kratter11}.
Then Eq. \ref{ppd_invisc_cond} relaxes to 
$\gamma,\, Q > 1$ in the outer disk. The condition on $\gamma$
is then guaranteed. On the other hand, numerical 
simulations of gravito-turbulence show that $1\lesssim Q \lesssim 2$
\citep{gammie01,rice11}, and the second inequality is generally
satisfied.   
%KMK
Taken together, this suggests that in the outer regions of a realistic
PPD, cooling %by itself may not be not responsible 
may not be the primary cause for a secondary 
instability of a gravito-turbulent disk, leading to
fragmentation. This leaves viscous GI as the only possible culprit
within our framework, as we illustrate below.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.0cm]{figures/ppd_2d_growth}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.cm]{figures/ppd_2d_maxk}
  \caption{Black lines show growth timescales (top) of the most
    unstable wavenumber (bottom) for viscous, 
    self-gravitational modes in the 2D PPD model shown in
    Fig. \ref{rafikov_model}. Blue curves are for the same disk
    model but with a lower accretion rate. Solid curves are obtained
    numerically from Eq. \ref{thindisk}, and dotted curves 
    are analytic results based on Eq. \ref{gammie_smallk}. 
    \label{rafikov_growth}}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.0cm]{figures/ppd_3d_rates}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.8cm]{figures/ppd_3d_maxk}
  \caption{Growth timescales (top) of the most unstable
    wavenumber (bottom) for viscous GI in a 3D PPD with radiative
    diffusion  (black lines). Asterisks are corresponding results
    obtained from the corresponding 2D problem with softened gravity.   
    % The black curves are obtained
  %  from the 3D eigenvalue problem with radiative diffusion,
   % applicable to optically-thick regions ($R\lesssim 60$AU); 
   % while the green curves are obtained from the 2D corresponding
   % model, which is applicable to both optically-thick and
   % optically-thin regimes.  
    \label{rafikov_growth3d}}
\end{figure}

>>>>

PPDs become irradiation-dominated at large distances from the star,
where $T\to\tirr$ and $\theta\to 1$ {\bf \citep{chiang97,dalessio97,kratter11}}.
Then Eq. \ref{ppd_invisc_cond} relaxes to 
$\gamma,\, Q > 1$ in the outer disk. The condition on $\gamma$
is then guaranteed. On the other hand, numerical 
simulations of gravito-turbulence show that $1\lesssim Q \lesssim 2$
\citep{gammie01,rice11}, and the second inequality is generally
satisfied.   
%KMK
Taken together, this suggests that in the outer regions of a realistic
PPD, cooling %by itself may not be not responsible 
may not be the primary cause for a secondary 
instability of a gravito-turbulent disk, leading to
fragmentation. This leaves viscous GI as the only possible culprit
within our framework, as we illustrate below.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.0cm]{figures/ppd_2d_growth}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.cm]{figures/ppd_2d_maxk}
  \caption{Black lines show growth timescales (top) of the most
    unstable wavenumber (bottom) for viscous, 
    self-gravitational modes in the 2D PPD model shown in
    Fig. \ref{rafikov_model}. Blue curves are for the same disk
    model but with a lower accretion rate. Solid curves are obtained
    numerically from Eq. \ref{thindisk}, and dotted curves 
    are analytic results based on Eq. \ref{gammie_smallk}. 
    {\bf 
      For both accretion rates the cooling time drops below $3/\Omega$
      beyond the vertical dashed line, but for the  
      $\dot{M}=10^{-6}M_\sun\mathrm{yr}^{-1}$ disk $\alpha > 0.1$
      beyond this radius, while $\alpha\lesssim
      0.03$  for the
      $\dot{M}=10^{-7}M_\sun\mathrm{yr}^{-1}$. 
    }
    \label{rafikov_growth}}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.0cm]{figures/ppd_3d_rates}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.8cm]{figures/ppd_3d_maxk}
  \caption{Growth timescales (top) of the most unstable
    wavenumber (bottom) for viscous GI in a 3D PPD with radiative
    diffusion  (black lines). Asterisks are corresponding results
    obtained from the corresponding 2D problem with softened gravity.   
    {\bf
      Beyond the vertical dashed line $\alpha>0.1$, which corresponds to
      cooling times $t_\mathrm{c}\lesssim 4/\Omega$.  
    }
    % The black curves are obtained
  %  from the 3D eigenvalue problem with radiative diffusion,
   % applicable to optically-thick regions ($R\lesssim 60$AU); 
   % while the green curves are obtained from the 2D corresponding
   % model, which is applicable to both optically-thick and
   % optically-thin regimes.  
    \label{rafikov_growth3d}}
\end{figure}

### 2016-03-24 paper/paper3d.tex
rough revision done

\subsection{Viscosity-driven gravitational instability}\label{visc_gi}
%there is also viscous damping which stabilizes
Viscous disks can also develop GI even when $Q>1$. This is because, as 
demonstrated below, viscosity removes rotational support against  
self-gravity for long-wavelength disturbances 
\citep{lynden-bell74,willerding92,gammie96}. A similar effect   
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. 
In fact, this
is a mechanism to enhance particle clumping for planetesimal
formation %frag of dust 
\citep{youdin11}. Therefore it is not unreasonable to expect analogous
fragmentation in gaseous disks due to viscosity.

@@

Consider an initially laminar, $Q\gg1$ disk, without external 
heating, as it cools. The disk temperature will decline
until $Q=O(1)$, whence non-axisymmetric modes grow and heat the disk
through the dissipation of spiral shocks \citep{cossins09}. 
%new rafikov 16  
This setup permits a quasi-steady, turbulent state with
$Q =O(1)$ in which cooling is balanced by shock heating to maintain
thermal equilibrium \citep{gammie01}. 
%If one represents this turbulent heating and transport {\bf by a viscosity} 
%heating by a viscous heating

>>>>

\subsection{Viscosity-driven gravitational instability}\label{visc_gi}
%there is also viscous damping which stabilizes
Viscous disks can also develop GI even when $Q>1$. This is because, as 
demonstrated below, viscosity removes rotational support against  
self-gravity for long-wavelength disturbances 
\citep{lynden-bell74,willerding92,gammie96}. A similar effect   
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag {\bf \citep{goodman00,ward00, takahashi14}}. 
In fact, this
is a mechanism to enhance particle clumping for planetesimal
formation %frag of dust 
{\bf \citep{youdin05,youdin11}}. Therefore it is not unreasonable to expect analogous
fragmentation in gaseous disks due to viscosity.

@@

Consider an initially laminar, $Q\gg1$ disk, without external 
heating, as it cools. The disk temperature will decline
until $Q=O(1)$, whence non-axisymmetric modes grow and heat the disk
through the dissipation of spiral shocks \citep{cossins09}. 
%new rafikov 16  
This setup permits a quasi-steady, turbulent state with
$Q =O(1)$ in which cooling is balanced by shock heating to maintain
thermal equilibrium {\bf \citep{gammie01,shi14}}. 
%If one represents this turbulent heating and transport {\bf by a viscosity} 
%heating by a viscous heating

### 2016-03-24 paper/results2d.tex
rough revision done

\subsubsection{$\theta = 1$}\label{theta1}
Standard beta cooling with $\theta=1$ corresponds to 
`thermal relaxation': the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case no
additional heat source need be invoked to define an inviscid steady state. From
Eq. \ref{stable_condition}, the instability condition is $Q<1$. %no
                                %beta dependence
This is the same as the classic Toomre condition for an isothermal
disk (which may be obtained from Eq. \ref{inviscid}  by taking
$|\beta S|\to 0$). In this respect, a 
fully irradiated disk, in which the equilibrium temperature is set
externally, behaves isothermally regardless of the cooling time.

@@

%{\bf maybe move following to discussion}
Numerical simulations of gravito-turbulent disks show there is a
maximum $\alpha$ ($\sim 0.06$) that can be sustained before
fragmentation \citep{rice05}. We can interpret this result in our
linear framework. 
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$ (second case in
Eq. \ref{high_visc}). Fig. \ref{gammie_maxrate_plot}  
shows that such a disk would be dynamically unstable with 
growth rate $s = O(\Omega)$. This is due to the direct effect of viscous 
stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
and hence highly turbulent, self-gravitating disks to persist beyond
dynamical timescales.
%KMK
We might interpret fragmentation {\bf as a reflection of} instability
of the highly viscous state. This is consistent with previous numerical
simulations performed by \cite{lodato05}. 
%Case (\ref{gvisc}) is consistent with 
\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    1.0cm]{figures/result2d_gvisc_kmax}
  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
    viscous GI as a function of cooling time for the
    case shown in the top panel of Fig. \ref{gammie_rate_plot}. Black curves 
    computed from the dispersion relation (Eq. \ref{thindisk}), and
    the orange curves are estimates based on
    Eq. \ref{gammie_smallk}---\ref{gammie_bigk}. 
    The dashed
    line marks the region with $\alpha > 1$, and the 
    dashed-dot line marks the region where growth timescales exceed
    the dynamical time.   
    \label{gammie_maxrate_plot}}
\end{figure}

>>>>

\subsubsection{$\theta = 1$}\label{theta1}
Standard beta cooling with $\theta=1$ corresponds to 
`thermal relaxation': the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case no
additional heat source need be invoked to define an inviscid steady state. From
Eq. \ref{stable_condition}, the instability condition is $Q<1$. %no
                                %beta dependence
This is the same as the classic Toomre condition for an isothermal
disk (which may be obtained from Eq. \ref{inviscid}  by taking
$|\beta S|\to 0$). In this respect, a 
fully irradiated disk, in which the equilibrium temperature is set
externally, behaves isothermally regardless of the cooling time 
{\bf \citep{gammie01,johnson03}}.

@@

%{\bf maybe move following to discussion}
Numerical simulations of gravito-turbulent disks show there is a
maximum $\alpha$ ($\sim 0.06$) that can be sustained before
fragmentation \citep{rice05}. We can interpret this result in our
linear framework. 
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$ (second case in
Eq. \ref{high_visc}). Fig. \ref{gammie_maxrate_plot}  
shows that such a disk would be dynamically unstable with 
growth rate $s = O(\Omega)$. This is due to the direct effect of viscous 
stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
and hence highly turbulent, self-gravitating disks to persist beyond
dynamical timescales.
%KMK
We might interpret fragmentation {\bf as an} instability
of the highly viscous state {\bf (see \S\ref{gi_of_gi})}. This is
consistent with previous numerical 
simulations performed by \cite{lodato05}. 
%Case (\ref{gvisc}) is consistent with 
\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    1.0cm]{figures/result2d_gvisc_kmax}
  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
    viscous GI as a function of cooling time for the
    case shown in the top panel of Fig. \ref{gammie_rate_plot}. Black curves 
    computed from the dispersion relation (Eq. \ref{thindisk}), and
    the orange curves are estimates based on
    Eq. \ref{gammie_smallk}---\ref{gammie_bigk}. 
    The dashed
    line marks the region with $\alpha > 1$, and the 
    dashed-dot line marks the region where growth timescales exceed
    the dynamical time.   
    \label{gammie_maxrate_plot}}
\end{figure}

### 2016-03-24 paper/summary.tex
rough revision done

More recent simulations have raised the issue of numerical
convergence. %KMK 
\cite{meru11} found that better resolved disks fragmented at longer cooling times.
Follow-up studies attributed at least some of this effect to decreasing numerical viscous
heating at higher resolution, which helps fragmentation
\citep{lodato11,meru12}. 
We can expect this if numerical viscosity
contributes to an effective irradiation, %then we do expect a disk can be
%destabilized at longer cooling times with increasing resolution, 
because then perturbations can cool to lower temperatures with
increasing resolution, see Fig. \ref{invisc_theta}.

 \cite{paardekooper12} also found that fragmentation can occur 
for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate/viscosity below  which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing $\tcool$ (decreasing $\alpha$), implying that
instabilities, and hence fragmentation, simply take longer to develop
for slowly-cooled disks.         
%hopkins paper

>>>>

More recent simulations have raised the issue of numerical
convergence. %KMK 
\cite{meru11} found that better resolved disks fragmented at longer cooling times.
Follow-up studies attributed at least some of this effect to decreasing numerical viscous
heating at higher resolution, which helps fragmentation
\citep{lodato11,meru12}. 
We can expect this if numerical viscosity
contributes to an effective irradiation, %then we do expect a disk can be
%destabilized at longer cooling times with increasing resolution, 
because then perturbations can cool to lower temperatures with
increasing resolution, see Fig. \ref{invisc_theta}.  {\bf
  In global simulations, non-convergence has also been attributed to initial 
  conditions that lead to internal edges \citep{paardekooper11b}, but
  this cannot be modeled in our local setup. 
  \cite{rice12} point out that the standard implementation of beta-cooling in
  SPH simulations apply cooling on scales well-below the SPH smoothing
  lengths, and that this inconsistency may contribute to
  non-convergence. 
}

{\bf However,} \cite{paardekooper12} also found {\bf in local grid-based simulations} that fragmentation can occur 
for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate/viscosity below  which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing $\tcool$ (decreasing $\alpha$), implying that
instabilities, and hence fragmentation, simply take longer to develop
for slowly-cooled disks.         
%hopkins paper
